\vspace{1.0cm} \LARGE \bf Structured population models demystify MERS-CoV epidemiology

### 2017-06-20 mers-structure.tex
Comments and revisons through Results/Seasonality
This is excellent. Stylistic changes (like "figure 1" to "Figure 1") are attempts to move closer to eLife style.

% basic packages
\usepackage{color}
\usepackage{parskip}
\usepackage{float}

@@

Middle East Respiratory Syndrome coronavirus (MERS-CoV) is a zoonotic virus originating in camels that has been causing significant mortality and morbidity in humans in the Arabian Peninsula.
The epidemiology of the virus remains poorly understood, with hospital outbreaks, isolated cases with known exposure to camels and apparent community transmission occurring simultaneously.
Whilst traditional and seroepidemiological studies have been employed extensively throughout the epidemic, sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
Here we use existing MERS-CoV sequence data to explore the phylodynamics of the virus in two of its known major hosts, humans and camels.
We employ state-of-the-art structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as the virus' terminal host.
By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that MERS-CoV is unlikely to become endemic in humans and that human outbreaks in the Arabian peninsula for the most part are driven by seasonally varying zoonotic transfer of viruses from camels.

@@

\section*{Introduction}
Middle East Respiratory Syndrome Coronavirus (MERS-CoV) is a zoonotic infection of humans in the Arabian Peninsula originating from camels.
The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has gone on to cause more than 1900 infections and numerous deaths, according to the World Health Organization (WHO).
Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected patients, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
Whilst contact with camels is often reported, patients may not recall contact with livestock, suggesting an unobserved community contribution to the outbreak \citep{group_state_2013}.
Studies into MERS-CoV epidemiology in the past often relied on serology to identify factors associated with MERS-CoV exposure in potential risk groups \citep{reusken_occupational_2015,reusken_2013}.
Indeed, such data have been used to show high seroprevalence in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014} and evidence of contact with MERS-CoV in workers with occupational exposure to camels \citep{reusken_occupational_2015,muller_presence_2015} compared to the general population.
Traditional epidemiological approaches have also been used to look at distributions of case clusters through time, space and across hosts \citep{cauchemez_unraveling_2016}.

Although traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much remains opaque to such approaches.
The new era of genomic epidemiology, however, has repeatedly shown the utility of sequence data in outbreak scenarios \citep{quick_real-time_2016,arias_rapid_2016}, where data are relatively cheap to produce, can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
Often only sequence data can pinpoint sources of pathogens and discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
A large portion of MERS-CoV sequences come from hospital outbreaks, where sequence data have been used to determine whether introductions were of a single origin and homogenous or a result of increased incidence \citep{fagbo_molecular_2015}.
Similar studies on MERS-CoV have taken place at broader geographic scales, such as cities \citep{cotten_2013}.

It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO's guidelines.
Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional, and necessarily limited, approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of population structure \citep{vaughan_efficient_2014}.
Unlike discrete trait analyses (DTA, or ``mugration'' models) that treat a potential trait of interest as if it were a nucleotide site with an arbitrary number of states, population structure models explicitly estimate distinct branching rates in each deme separately.
By jointly estimating independent coalescence rates for MERS-CoV in humans and camels, as well as migration patterns between the two proposed demes, we show that long-term viral evolution of MERS-CoV occurs exclusively in camels.
Our results suggest that spillover events into humans are seasonal and might be associated with the calving season in camels.
Once human MERS-CoV infections are established, however, we find that MERS-CoV is poor at transmitting between humans.
Using approximate Bayesian computation we show that R$_{0}$ for MERS-CoV circulating in humans is much lower than the epidemic threshold of 1.0 and that conversely the virus has been introduced into humans hundreds of times.

@@

\subsection*{MERS-CoV is predominantly a camel virus}
The structured coalescent approach we employ \citep{vaughan_efficient_2014} identifies camels as a reservoir host where most of MERS-CoV evolution takes place (figure \ref{mcc}), whilst human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (figure \ref{exploded}).
Across 174 MERS-CoV genomes collected from humans, we detect a median of 56 individual camel to human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
Whilst we identify a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (figure \ref{mcc}).
We conclude that most human-to-camel migrations are induced by the migration rate prior, whilst some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where poor sampling of MERS-CoV genetic diversity from camels allows the model to consider humans as a potential alternative host.
The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
Our finding of negligible human-to-camel transmission is robust to choice of prior (figure \ref{prior}).

The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (figure \ref{mcc} and \ref{exploded}).
We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and robust to choice of prior.
We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to index cases that might be mild and ultimately not detected by surveillance or sequencing.

@@

\subsection*{Signals of MERS-CoV seasonality}
Unlike the most basic ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, an additional advantage structured population models have over less sophisticated approaches is the ability to recover probable distribution of migration times.
We use the posterior distribution of MERS-CoV introduction times for human clades observed in the MCC tree (figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels that leads to seasonal variation in force of infection in humans.
Strong influx of susceptibles and subequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology.

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_seasonality.png}
	\caption{\textbf{MERS-CoV seasonality.}
A) Violin plots coloured by month showing the 95\% highest posterior density interval for odds ratios of MERS-CoV introductions into human populations.
Posterior means are indicated with circles.
Evidence for increased or decreased risk (95\% HPD excludes 1.0) for introductions are indicated by black or white circles, respectively.
Grey area spanning October to February indicates the camel calving season.
B) Each clade entering humans is positioned based on the median introduction time, as recovered by structured coalescent analyses and coloured by time of year with height indicating number of descendent sequences recovered from human cases.
95\% highest posterior density intervals for zoonotic introduction of MERS-CoV into human populations observed in the MCC tree (figure \ref{mcc}) are indicated with coloured lines, coloured by median estimated introduction time.
The black dotted line indicates the joint probability density for introductions.
Grey areas spanning October to February indicate the reported camel calving season.
	}
	\label{seasonality}
\end{figure}

@@

We identify R$_{0}$ values below 1.0 as being more likely to generate MERS-CoV-like distributions of sequence clusters (middle panels in figure \ref{ABC}).
For increasing levels of bias mean R$_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
Whilst the 95\% percentiles for R$_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.} uniform sequencing efforts), we also note that increasing levels of bias are considerably better at generating MERS-CoV-like sequence clusters.
Under unbiased sequencing only 2282 (out of 363 000) simulations fit our observations in structured coalescent analyses of MERS-CoV which increase to 9763 and 9923 for bias levels of 2.0 and 3.0, respectively.
This is apparent in the heights of the histograms in figure \ref{ABC}.
Simulations with bias suggest that R$_{0}$ values could have ranges [0.605--0.870] (bias = 2) and [0.555--0.825] (bias = 3).

Lower values for R$_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans.
The median number of cross-species introductions observed in simulations matching MERS data without bias are 339 (95\% percentiles 255--431).
These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously seem to produce considerably more MERS-CoV-like data.
Our results suggest a large number of unobserved MERS index cases.
Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
In those sets of analyses R$_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (figure \ref{extra_cases}).
Overall our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated R$_{0}<1.0$ and sequence sampling wholly unrepresentative of viral diversity found in humans.
Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices appear sufficient to explain observed epidemiological patterns of MERS-CoV.

@@

Both tests can give spurious results in cases of extreme rate heterogeneity and sampling over time \citep{dudas_mers-cov_2016}, but both tests have not been reported to fail simultaneously.
PHI and 3Seq methods consistently identify most of the apparent `backbone' of the MERS-CoV phylogeny as encompassing sequences with evidence of recombination (figure \ref{recombination_tree}).
Neither method can identify where in the tree recombination occurred, but each full asterisk in figure \ref{recombination_tree} should be interpreted as the minimum partition of data that still captures both donor and recipient alleles involved in a recombination event.
This suggests a non-negligible contribution of recombination in shaping existing MERS-CoV diversity.
As before \citep{dudas_mers-cov_2016}, we show large numbers of homoplasies in MERS-CoV data (figure \ref{incompatibilities}) with some evidence of spatial clustering of such alleles.
Homoplasies present in multiple strains at a time, indicating ancestral events that have been successful.
Though the proportion of time that the virus is inferred to spend in camels alone may be sufficient to argue that camels are the host where MERS-CoV recombines, incidence of MERS-CoV is known to be much higher in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2014,ali_systematic_2017}.
This provides ideal conditions for co-infection, which is a pre-requisite for RNA virus recombination to occur.

Conversely, our results argue strongly that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
We find little evidence that recombination might be interfering with human outbreak cluster inference beyond what could be expected of phylogenetic noise (figure \ref{recombination}A).
And whilst we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (figure \ref{recombination}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (figure \ref{flower}).
Observed departures from strictly clonal evolution suggest that whilst recombination is a serious issue in MERS-CoV phylogenies, its effects on the human side of MERS outbreaks is minimal, as expected.
Overall we advise extreme caution when interpreting branches in MERS-CoV phylogenies that are suspected of having spent any time in camels, especially when attempting to identify origins of human cases.
Conversely, the largely clonal evolution of MERS-CoV in humans, despite limited diversity, should be more amenable to traditional phylogenetic approaches.

@@

Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in MERS-CoV relative genetic diversity (N$_{e}\tau$) over time (figure \ref{skygrid}).
However, we do note that estimates of relative genetic diversity are relatively low overall, and MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
This raises the possibility that MERS-CoV populations undergo lineage turnover at a rate not often seen in other livestock viruses (\textit{e.g.} swine influenza A viruses \citep{vijaykrishna_long-term_2011}).
Part of this could be caused by MERS-CoV ability to re-infect camels previously exposed to the virus \citep{ali_systematic_2017}, which may result in a process akin to antigenic drift, where the depletion of susceptible calves leads to ever increasing selective pressure for the virus to evolve antigenically.
Whilst this could be tested by looking for seasonally varying patterns in selection on MERS-CoV surface proteins, recombination would remain a key confounding problem.

@@

We showed that currently existing models of population structure \citep{vaughan_efficient_2014} can identify distinct demographic modes in MERS-CoV data, where viruses continuously circulating in camels repeatedly jump into humans and cause small outbreaks doomed to extinction (figures \ref{mcc}, \ref{exploded}).
This inference succeeds under different choices of priors for unknown demographic parameters (figure \ref{prior}) and in the presence of strong biases in sequence sampling schemes (figure \ref{ABC}).
From sequence data we identify at least 50 zoonotic introductions of MERS-CoV into humans from the reservoir (figure \ref{mcc}), from which we extrapolate that hundreds more such introductions must have taken place (figure \ref{ABC}).
We also looked at potential seasonality in MERS-CoV spillover into humans, which is an additional advantage of explicitly modeling population structure.
Our analyses indicated a period of three months where the odds of a sequenced spillover event are increased, with timing consistent with an enzootic amongst camel calves (figure \ref{seasonality}).
As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (figure \ref{ABC}).
Poor capacity of MERS-CoV to spread human-to-human explains why evidence of substantial community transmission has been lacking, as well as the absence of virus circulation outside of Arabian peninsula.
Perhaps a clearer indicator of this has been the lack of a global MERS pandemic, which would have certainly taken place had R$_{0}$ been above the epidemic threshold.

Despite higher resolution of inference offered by structured population models, especially when sequence data are plentiful, serious drawbacks exist.
As the representation of evolution drifts away from a strictly bifurcating tree model, the space over which phylogenetic inference needs to take place becomes increasingly complex.
Convergence difficulties are well known in rich models, such as ancestral recombination graphs, and multitype trees are no different.
Even with a moderate number of sequences (N=274) and the smallest possible number of demes (N=2) structured population model mixing is slow and problematic.
Some convergence issues (figure \ref{skygrid}) could be attributed to the presence of recombination in MERS-CoV data (figure \ref{recombination}), but active research into approximations to the structured coalescent \citep{maio_new_2015} seem to acknowledge both the utility and difficulties in applying such methods.
As computational infrastructure improves \citep{ayres_beagle:_2012}, it is hopefully only a matter of time before phylogenetic population structure models mature to the point where large datasets with multiple demes (\textit{e.g.} West African Ebola virus epidemic \citep{dudas_virus_2017}, global influenza virus circulation \citep{bedford_global_2015}) can be analysed trivially.

@@

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_exploded.png}
	\caption{\textbf{Evolutionary history of MERS-CoV partitioned between camels and humans.}
Same tree as shown in figure \ref{mcc}, but all contiguous stretches of MERS-CoV evolutionary history split by inferred host: camels (top) and humans (bottom).
This visualisation highlights the ephemeral nature of MERS-CoV outbreaks in humans, compared to continuous circulation of the virus in camels.
	}
	\label{exploded}
\end{figure}

>>>>

% basic packages
\usepackage{color}
\usepackage{parskip}
\usepackage{float}
\usepackage{url}
\urlstyle{same}

@@

Middle East Respiratory Syndrome coronavirus (MERS-CoV) is a zoonotic virus originating in camels that has been causing significant mortality and morbidity in humans in the Arabian Peninsula.
The epidemiology of the virus remains poorly understood, with hospital outbreaks, isolated cases with known exposure to camels and apparent community transmission occurring simultaneously.
Whilst traditional and seroepidemiological studies have been employed extensively throughout the epidemic, viral sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
Here we use existing MERS-CoV sequence data to explore the phylodynamics of the virus in two of its known major hosts, humans and camels.
We employ state-of-the-art structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as the virus' terminal host.
\tbc{Fields' Virology textbook defines `terminal host' as `one that is not involved in maintaining the virus, such infections rarely result in transmission to other human or non-human hosts'. I think you want a different word here. Maybe `transient host'?}
By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that MERS-CoV is unlikely to become endemic in humans and that human outbreaks in the Arabian peninsula for the most part are driven by seasonally varying zoonotic transfer of viruses from camels.
\tbc{Powerful and succient abstract. Great.}

@@

\section*{Introduction}
Middle East Respiratory Syndrome Coronavirus (MERS-CoV) is a zoonotic infection of humans in the Arabian Peninsula originating from camels.
The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has gone on to cause more than 2000  infections and over 700 deaths, according to the World Health Organization (WHO) \citep{mersfactsheet}.
Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected individuals, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
Whilst contact with camels is often reported, patients may not recall contact with livestock, suggesting an unobserved community contribution to the outbreak \citep{group_state_2013}.
Studies into MERS-CoV epidemiology in the past often relied on serology to identify factors associated with MERS-CoV exposure in potential risk groups \citep{reusken_occupational_2015,reusken_2013}.
Indeed, such data have been used to show high seroprevalence in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014} and evidence of contact with MERS-CoV in workers with occupational exposure to camels \citep{reusken_occupational_2015,muller_presence_2015}.
Epidemiological modeling approaches have been used to look at traditional data sources of case clusters through time, space and across hosts \citep{cauchemez_unraveling_2016}.

Although traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much, inevitably, remains opaque to such approaches.
The new era of genomic epidemiology, however, has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{quick_real-time_2016,arias_rapid_2016}, where data are relatively cheap to produce, can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
Often only sequence data can pinpoint sources of pathogens and discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
A large portion of MERS-CoV sequences come from hospital outbreaks, where sequence data have been used to determine whether infections were isolated introductions or were part of a larger hospital-associated outbreak \citep{fagbo_molecular_2015}.
Similar studies on MERS-CoV have taken place at broader geographic scales, such as cities \citep{cotten_2013}.

It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines.
Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional, and necessarily limited, approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of metapopulation structure and migration between discrete sub-populations, referred to here as demes \citep{vaughan_efficient_2014}.
Unlike discrete trait analyses (DTA, or ``mugration'' models) that treat a potential trait of interest (such as host state) as if it were a nucleotide site with an arbitrary number of states, population structure models explicitly incorporate distinct sampling patterns and population dynamics across demes.
By estimating independent coalescence rates for MERS-CoV in humans and camels, as well as migration patterns between the two demes, we show that long-term viral evolution of MERS-CoV occurs exclusively in camels.
Our results suggest that spillover events into humans are seasonal and might be associated with the calving season in camels.
Once human MERS-CoV infections are established, however, we find that MERS-CoV is poor at transmitting between humans.
Using approximate Bayesian computation we show that $R_{0}$ for MERS-CoV circulating in humans is much lower than the epidemic threshold of 1.0 and that correspondingly the virus has been introduced into humans hundreds of times.

@@

\subsection*{MERS-CoV is predominantly a camel virus}
The structured coalescent approach we employ \citep{vaughan_efficient_2014} identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), whilst human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure \ref{exploded}).
Across 174 MERS-CoV genomes collected from humans, we detect a median of 56 individual camel-to-human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
Whilst we identify a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure \ref{mcc}).
We conclude that most human-to-camel migrations are induced by the migration rate prior, whilst some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where poor sampling of MERS-CoV genetic diversity from camels allows the model to consider humans as a potential alternative host.
The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
Our finding of negligible human-to-camel transmission is robust to choice of prior (Figure \ref{prior}).

The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure \ref{mcc} and \ref{exploded}).
We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to index cases that might be mild and ultimately not detected by surveillance or sequencing.

@@

\subsection*{Signals of MERS-CoV seasonality}
Unlike the most basic ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, an additional advantage structured population models have over less sophisticated approaches is the ability to estimate migration times.
We use the posterior distribution of MERS-CoV introduction times for human clades observed in the MCC tree (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (Figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels that leads to seasonal variation in force of infection in humans.
Strong influx of susceptibles and subequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology.

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_seasonality.png}
	\caption{\textbf{MERS-CoV seasonality.}
A) Violin plots coloured by month showing the 95\% highest posterior density interval for odds ratios of MERS-CoV introductions into human populations.
Posterior means are indicated with circles.
Evidence for increased or decreased risk (95\% HPD excludes 1.0) for introductions are indicated by black or white circles, respectively.
Grey area spanning October to February indicates the camel calving season.
B) Each clade entering humans is positioned based on the median introduction time, as recovered by structured coalescent analyses and coloured by time of year with height indicating number of descendent sequences recovered from human cases.
95\% highest posterior density intervals for zoonotic introduction of MERS-CoV into human populations observed in the MCC tree (Figure \ref{mcc}) are indicated with coloured lines, coloured by median estimated introduction time.
The black dotted line indicates the joint probability density for introductions.
Grey areas spanning October to February indicate the reported camel calving season.
	}
	\label{seasonality}
\end{figure}

@@

We identify R$_{0}$ values below 1.0 as being more likely to generate MERS-CoV-like distributions of sequence clusters (middle panels in Figure \ref{ABC}).
For increasing levels of bias mean R$_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
Whilst the 95\% percentiles for R$_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.} uniform sequencing efforts), we also note that increasing levels of bias are considerably better at generating MERS-CoV-like sequence clusters.
Under unbiased sequencing only 2282 (out of 363 000) simulations fit our observations in structured coalescent analyses of MERS-CoV which increase to 9763 and 9923 for bias levels of 2.0 and 3.0, respectively.
This is apparent in the heights of the histograms in Figure \ref{ABC}.
Simulations with bias suggest that R$_{0}$ values could have ranges [0.605--0.870] (bias = 2) and [0.555--0.825] (bias = 3).

Lower values for R$_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans.
The median number of cross-species introductions observed in simulations matching MERS data without bias are 339 (95\% percentiles 255--431).
These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously seem to produce considerably more MERS-CoV-like data.
Our results suggest a large number of unobserved MERS index cases.
Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
In those sets of analyses R$_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{extra_cases}).
Overall our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated R$_{0}<1.0$ and sequence sampling wholly unrepresentative of viral diversity found in humans.
Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices appear sufficient to explain observed epidemiological patterns of MERS-CoV.

@@

Both tests can give spurious results in cases of extreme rate heterogeneity and sampling over time \citep{dudas_mers-cov_2016}, but both tests have not been reported to fail simultaneously.
PHI and 3Seq methods consistently identify most of the apparent `backbone' of the MERS-CoV phylogeny as encompassing sequences with evidence of recombination (Figure \ref{recombination_tree}).
Neither method can identify where in the tree recombination occurred, but each full asterisk in Figure \ref{recombination_tree} should be interpreted as the minimum partition of data that still captures both donor and recipient alleles involved in a recombination event.
This suggests a non-negligible contribution of recombination in shaping existing MERS-CoV diversity.
As before \citep{dudas_mers-cov_2016}, we show large numbers of homoplasies in MERS-CoV data (Figure \ref{incompatibilities}) with some evidence of spatial clustering of such alleles.
Homoplasies present in multiple strains at a time, indicating ancestral events that have been successful.
Though the proportion of time that the virus is inferred to spend in camels alone may be sufficient to argue that camels are the host where MERS-CoV recombines, incidence of MERS-CoV is known to be much higher in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2014,ali_systematic_2017}.
This provides ideal conditions for co-infection, which is a pre-requisite for RNA virus recombination to occur.

Conversely, our results argue strongly that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
We find little evidence that recombination might be interfering with human outbreak cluster inference beyond what could be expected of phylogenetic noise (Figure \ref{recombination}A).
And whilst we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombination}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
Observed departures from strictly clonal evolution suggest that whilst recombination is a serious issue in MERS-CoV phylogenies, its effects on the human side of MERS outbreaks is minimal, as expected.
Overall we advise extreme caution when interpreting branches in MERS-CoV phylogenies that are suspected of having spent any time in camels, especially when attempting to identify origins of human cases.
Conversely, the largely clonal evolution of MERS-CoV in humans, despite limited diversity, should be more amenable to traditional phylogenetic approaches.

@@

Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in MERS-CoV relative genetic diversity (N$_{e}\tau$) over time (Figure \ref{skygrid}).
However, we do note that estimates of relative genetic diversity are relatively low overall, and MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
This raises the possibility that MERS-CoV populations undergo lineage turnover at a rate not often seen in other livestock viruses (\textit{e.g.} swine influenza A viruses \citep{vijaykrishna_long-term_2011}).
Part of this could be caused by MERS-CoV ability to re-infect camels previously exposed to the virus \citep{ali_systematic_2017}, which may result in a process akin to antigenic drift, where the depletion of susceptible calves leads to ever increasing selective pressure for the virus to evolve antigenically.
Whilst this could be tested by looking for seasonally varying patterns in selection on MERS-CoV surface proteins, recombination would remain a key confounding problem.

@@

We showed that currently existing models of population structure \citep{vaughan_efficient_2014} can identify distinct demographic modes in MERS-CoV data, where viruses continuously circulating in camels repeatedly jump into humans and cause small outbreaks doomed to extinction (Figures \ref{mcc}, \ref{exploded}).
This inference succeeds under different choices of priors for unknown demographic parameters (Figure \ref{prior}) and in the presence of strong biases in sequence sampling schemes (Figure \ref{ABC}).
From sequence data we identify at least 50 zoonotic introductions of MERS-CoV into humans from the reservoir (Figure \ref{mcc}), from which we extrapolate that hundreds more such introductions must have taken place (Figure \ref{ABC}).
We also looked at potential seasonality in MERS-CoV spillover into humans, which is an additional advantage of explicitly modeling population structure.
Our analyses indicated a period of three months where the odds of a sequenced spillover event are increased, with timing consistent with an enzootic amongst camel calves (Figure \ref{seasonality}).
As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (Figure \ref{ABC}).
Poor capacity of MERS-CoV to spread human-to-human explains why evidence of substantial community transmission has been lacking, as well as the absence of virus circulation outside of Arabian peninsula.
Perhaps a clearer indicator of this has been the lack of a global MERS pandemic, which would have certainly taken place had R$_{0}$ been above the epidemic threshold.

Despite higher resolution of inference offered by structured population models, especially when sequence data are plentiful, serious drawbacks exist.
As the representation of evolution drifts away from a strictly bifurcating tree model, the space over which phylogenetic inference needs to take place becomes increasingly complex.
Convergence difficulties are well known in rich models, such as ancestral recombination graphs, and multitype trees are no different.
Even with a moderate number of sequences (N=274) and the smallest possible number of demes (N=2) structured population model mixing is slow and problematic.
Some convergence issues (Figure \ref{skygrid}) could be attributed to the presence of recombination in MERS-CoV data (Figure \ref{recombination}), but active research into approximations to the structured coalescent \citep{maio_new_2015} seem to acknowledge both the utility and difficulties in applying such methods.
As computational infrastructure improves \citep{ayres_beagle:_2012}, it is hopefully only a matter of time before phylogenetic population structure models mature to the point where large datasets with multiple demes (\textit{e.g.} West African Ebola virus epidemic \citep{dudas_virus_2017}, global influenza virus circulation \citep{bedford_global_2015}) can be analysed trivially.

@@

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_exploded.png}
	\caption{\textbf{Evolutionary history of MERS-CoV partitioned between camels and humans.}
Same tree as shown in Figure \ref{mcc}, but all contiguous stretches of MERS-CoV evolutionary history split by inferred host: camels (top) and humans (bottom).
This visualisation highlights the ephemeral nature of MERS-CoV outbreaks in humans, compared to continuous circulation of the virus in camels.
	}
	\label{exploded}
\end{figure}

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_prior.png}
	\caption{\textbf{Result are robust to prior choice.}
Negligible flow of MERS-CoV lineages from humans into camels is recovered regardless of prior choice.
Plots on the left show the 95\% highest posterior density for the estimated migration rate from the human deme into the camel deme (left) and \textit{vice versa} (right).
Dotted lines indicate exponential priors specified for migration rates, with mean 1.0 (top) or 10.0 (bottom).
\tbc{I think this would be more clear if the left and right sides shared the same x-axis. As it stands it looks at first glance like there's different priors for camel-to-human compared to human-to-camel. It's okay if the blue human-to-camel rates gets smushed up against the left of the plot.}
	}
	\label{prior}
\end{figure}
